dyngen: benchmarking with in silico single cells

### 2019-12-19 manuscript.Rmd
update manuscript

**Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.  
**Results:** We developed dyngen, a multi-modality simulator of single cells. In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network. By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures. We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments.  
**Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.

@@

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions, thereby limiting the applicability of each generator.

@@

library(kableExtra)
tab %>%
  select(-id, -type, -model, -citation, -package) %>%
  spread(plabel, check) %>%
  knitr::kable(escape = FALSE) %>%
  kableExtra::row_spec(0, angle = 0)
```

@@

Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases. In practice, dyngen has already successfully been used to evaluate trajectory inference[@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression[@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference[@pratapa_benchmarkingalgorithmsgene_2019] methods.
To demonstrate this point even further, we apply dyngen on several promising novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.

@@

Trajectory alignment is a technique used to discover the similarities and differences between the developmental trajectories of cells. 
This technique has been used to compare gene expression kinetics resulting from different biological processes[@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development[@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors[@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis[@alpert_alignmentsinglecelltrajectories_2018].

Unfortunately, not many methods to align trajectories exist. DTW (Dynamic Time Warping)[@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories. DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between two sequences that indicates which sequence must be dilated or contracted to best match the other one.

dyngen can be used to evaluate trajectory alignment techniques. We generated two datasets and permuted each of those datasets in five ways: we knocked down the transcription factors belonging to the end of the second milestone, we set 10% or 20% of the count matrix to zero, and we performed the combination of those techniques. This results in 12 datasets, each containing a slightly different trajectory.
We compared DTW when using smoothed pseudocells (with a gaussian kernel) and using just all cells. For each combination of trajectories, we calculated the DTW alignment. In figure x, the ratio between one trajectory with a certain permutation, and the other trajectory with the same permutation is shown. We expect this ratio to be low, as we want the alignment to be able to distinguish between different trajectories. We especially expect the smoothed versions to be able to overcome the extra 10% or 20% noise.
We performed this experiment 10 times.

In Figure \ref{fig:traj_align}, we can see that DTW with gaussian smoothing cannot distinguish between trajectory 1 and trajectory 2. When using the cells themselves, we find that they can clearly distinguish between trajectory 1 and 2 when noise is added, but not when the trajectory is limited to milestone C.

<!--
\begin{figure}[htb!]
    \centering
    \includegraphics[width=.8\linewidth]{fig/traj_alignment/tares.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
    \label{fig:traj_align}
\end{figure}
-->

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.8\textwidth]{fig/backbone_lego}
    \caption{
      \textbf{Backbone lego.}
    }
    \label{fig:backbone_lego}
\end{figure}

@@

**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone. These interactions inherit the effect, strength, and cooperativity parameters from the interactions in the backbone. A TF can only be regulated by other TFs or itself.

@@

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\hill}[1]{\text{hill}_{#1}}
\newcommand{\coop}[1]{\text{co}_{#1}}
\newcommand{\dis}[1]{\text{dis}_{#1}}
\newcommand{\buf}[1]{\chi_{#1}}
\newcommand{\ba}[1]{\text{ba}_{#1}}

@@

\begin{table}[ht]
    \caption{
      \textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mature mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
    } \label{tab:reaction_def}
    \centering
    \begin{tabular}{|lcc|}
        \hline
        Reaction & Effect & Propensity \\ \hline \hline
        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \coop{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\coop G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
        Pre-mRNA degradation & $\w G \rightarrow \emptyset$ & $\wdr G \times \w G$ \\
        Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
        Mature mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
        Translation & $\x G \rightarrow \x G + \y G$ & $\ypr G \times \x G$ \\
        Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
    \end{tabular}
\end{table}

\begin{table}[ht]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_params}
    \centering
    \begin{tabular}{|lrl|}
        \hline
        Parameter & Symbol & Definition \\ \hline \hline
        Transcription rate & $\wpr{G}$ & $\in N(50, 10),\ \geq 10$ \\
        Splicing rate & $\wsr G$ & $\in N(5, 1),\ \geq 1$ \\
        Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 1$ \\
        Pre-mRNA half-life & $\whl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
        Mature mRNA half-life & $\xhl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
        Protein half-life rate & $\yhl G$ & $\in N(0.25, 0.05),\ \geq 0.1$ \\
        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
        Pre-mRNA degradation rate & $\wdr G$ & $= \ln(2)\ /\ \whl G$ \\
        Mature mRNA degradation rate & $\xdr G$ & $= \ln(2)\ /\ \xhl G$ \\
        Protein degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \wsr H \times \ypr H}{(\wdr H + \wsr H) \times \xdr H \times \ydr H}$ \\
        Binding & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

@@

The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$. Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors. The following subsections explain and define the propensity function when $N=1$, $N=2$, and finally for an arbitrary $N$.

@@

LIONESS[@@kuijjer_estimatingsamplespecificregulatory_2019; kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer casewise GRNs. To do so, first the Pearson correlation is calculated between regulators and targets for all samples. Next, the Pearson correlation is again calculated for all samples excluding one sample. The difference between the two correlation matrices is considered a casewise GRN for that particular profile. This process is repeated for all profiles, resulting in a casewise GRN.

@@

The AUROC and AUPR metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN. To compare a predicted casewise GRN with the ground-truth casewise GRN, the top 10'000 interactions per cell were retained. For each cell-specific network, the AUROC and AUPR were calculated.
## Comparison of RNA velocity methods {#sec:dyngen-velcompare}

@@

>>>>

**Purpose:** When developing new types of tools for single-cell analyses, there is often a lack of datasets on which to quantitatively assess the performance.A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.  
**Results:** We developed dyngen, a multi-modality simulator of single cells. In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a predefined gene regulatory network.  We used dyngen to benchmark three upcoming methods for single-cell analyses: RNA velocity, casewise network inference and trajectory alignment methods. The simulator is easily extendible for adding new modalities or experimental procedures. We demonstrate dyngen's flexibility by applying it to benchmark RNA velocity, casewise network inference and trajectory alignment methods. 
**Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.

@@

Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data. Generators of scRNA-seq data (e.g. splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
However, a limitation of existing scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions., thereby limiting the applicability of each generator.

@@

library(kableExtra)
tab %>%
  select(-id, -type, -model, -citation, -package) %>%
  spread(plabel, check) %>%
  knitr::kable(escape = FALSE) %>%
  kableExtra::row_spec(0, angle = 0)
```
method
model
modalities
(possible) applications
splatter[@zappia_splattersimulationsinglecell_2017]
Several distributional models with parameters that allow the generation of datasets by sampling from distributions
Cell types, linear trajectories and batch effects can be simulated by changing the parameters of certain genes
mRNA expression
Clustering
Linear trajectory inference
Batch effect correction
Differential expression
powsimR[@vieth_powsimrpoweranalysis_2017]
Negative binomial or zero-inflated negative binomial.
Cell types can be simulated by changing the parameters of this distribution
mRNA expression
Differential expression
Clustering
PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018]
(1) Generation of a random tree
(2) The expression of each gene is seen as a sum of contributions from different expression programs. The activity of each program depends on pseudotime.
(3) Generation of cells and expression by random sampling along the trajectory, and from a negative binomial distribution
mRNA expression
Linear and branching trajectory inference
SymSim[@zhang_simulatingmultiplefaceted_2019]
Kinetic model of each gene, in which it can be on or off, and where the mRNA can be degraded. The parameters of this model depend on a gene effect and cell effect, each of these effects represents a vector in an n-dimensional space (e.g. 10). The cell effect vectors are sampled in such a way that cells that have to be more similar (according to a tree provided by the user), will also have a more similar vector.
mRNA expression
promoter activity
Clustering
Linear, branching and disconnected trajectory inference

dyngen
Simulation of a thermodynamic model of gene regulation, in which a set of transcription factors can bind the promoter and can have varying effects on the activity of the promoter. Clusters or trajectories are simulated by specifying a regulatory network that will induce a certain diversification of expression. Expression is simulated using a single-molecule model, after which individual UMIs are simulated using a multinomial distribution. Apart from gene regulation, splicing and protein production are also simulated.
Protein expression
mRNA expression
pre-mRNA expression
promoter activity
lineage state
reaction activity
Clustering
Linear, branching and disconnected trajectory inference
Cellwise network inference
Batch effect correction
RNA velocity
Trajectory alignment

@@

Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases. In practice, dyngen has already successfully been used to evaluate trajectory inference[@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression[@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference[@pratapa_benchmarkingalgorithmsgene_2019] methods.
To demonstrate this point even further, we apply dyngen on several promising novel computational approaches for which quantitative assessment of the performance was hitherto lacking.

@@

Trajectory alignment allows studying the differences between multiple trajectories that are mostly similar. For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage branch.

However, aligning trajectories becomes exponentially more difficult as the complexity of compared trajectories increases. For this reason, trajectory alignment remains a mostly unexplored territory within single-cell omics.

Dynamic Time Warping (DTW)[@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories. DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between the sequences that indicates which sequence must be dilated or contracted to best match the other one. DTW can also be used in combination with smoothed pseudocells[@alpert_alignmentsinglecelltrajectories_2018].

Trajectory alignment, using DTW, has been used to compare gene expression kinetics resulting from different biological processes[@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development[@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors[@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis[@alpert_alignmentsinglecelltrajectories_2018].

dyngen can be used to evaluate trajectory alignment techniques. We look how well DTW can continue to distinguish between different trajectories when an increasing amount of noise is added. In Figure \ref{fig:traj_align} we can see that when noise increases, DTW can no longer distinguish trajectories clearly, and that in this case, there is no difference whether smoothed pseudocells are used or when actual cells and their gene expression are used.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=.8\linewidth]{fig/traj_alignment/tares.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
  \textbf{A: Example of trajectory alignment using DTW. The segments between the two trajectories show which cells are aligned.}
\textbf{B: Shows the discrimination power of DTW decreasing as the amount of noise added increases.}
    \label{fig:traj_align}
\end{figure}

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.8\textwidth]{fig/backbone_lego}
    \caption{
      \textbf{Reusable regulatory building blocks that can be used to construct GRNs that result in dynamic processes following an expected pattern.}
      \textbf{A:} Each building block contains genes which serve as inputs that can be regulated by external transcription factors. The connections inside the building block transform the input signal in a particular way and has one or more output interactions of which the targets are not yet determined.
      \textbf{B:} By connecting multiple building blocks together, a backbone of regulatory modules can be formed.
      \textbf{C:} dyngen offers several variations of linear building blocks which allow for differing regulatory mechanisms.
      \textbf{D:} A branching module will upregulate the expression of just one of the output interactions. Two variations are given of which the simple variant is more prone to reprogramming events in comparison to the robust branching module.
      \textbf{E,F:} These components allow to specify the beginning (no inputs) and end (no outputs) of the backbone.
    }
    \label{fig:backbone_lego}
\end{figure}

@@

**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone. These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone. A TF can only be regulated by other TFs or itself.

@@

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\hill}[1]{\text{hill}_{#1}}
\newcommand{\ind}[1]{\text{ind}_{#1}}
\newcommand{\dis}[1]{\text{dis}_{#1}}
\newcommand{\buf}[1]{\chi_{#1}}
\newcommand{\ba}[1]{\text{bas}_{#1}}

@@

\begin{table}[ht]
    \caption{
      \textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mature mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
    } \label{tab:reaction_def}
    \centering
    \begin{tabular}{|lcc|}
        \hline
        Reaction & Effect & Propensity \\ \hline \hline
        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
        Pre-mRNA degradation & $\w G \rightarrow \emptyset$ & $\wdr G \times \w G$ \\
        Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
        Mature mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
        Translation & $\x G \rightarrow \x G + \y G$ & $\ypr G \times \x G$ \\
        Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
    \end{tabular}
\end{table}

\begin{table}[ht]
    \caption{
      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
    } \label{tab:reaction_params}
    \centering
    \begin{tabular}{|lrl|}
        \hline
        Parameter & Symbol & Definition \\ \hline \hline
        Transcription rate & $\wpr{G}$ & $\in N(50, 10),\ \geq 10$ \\
        Splicing rate & $\wsr G$ & $\in N(5, 1),\ \geq 1$ \\
        Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 1$ \\
        Pre-mRNA half-life & $\whl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
        Mature mRNA half-life & $\xhl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
        Protein half-life rate & $\yhl G$ & $\in N(0.25, 0.05),\ \geq 0.1$ \\
        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
        Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
        Pre-mRNA degradation rate & $\wdr G$ & $= \ln(2)\ /\ \whl G$ \\
        Mature mRNA degradation rate & $\xdr G$ & $= \ln(2)\ /\ \xhl G$ \\
        Protein degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \wsr H \times \ypr H}{(\wdr H + \wsr H) \times \xdr H \times \ydr H}$ \\
        Binding & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
    \end{tabular}
\end{table}

@@

The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation[@schilstra_biologicgeneexpression_2008needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$. Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors. The following subsections explain and define the propensity function when $N=1$, $N=2$, and finally for an arbitrary $N$.

@@

LIONESS[@@kuijjer_estimatingsamplespecificregulatory_2019; @kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer casewise GRNs. To do so, first the Pearson correlation is calculated between regulators and targets for all samples. Next, the Pearson correlation is again calculated for all samples excluding one sample. The difference between the two correlation matrices is considered a casewise GRN for that particular profile. This process is repeated for all profiles, resulting in a casewise GRN.

@@

The AUROC and AUPR metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN. To compare a predicted casewise GRN with the ground-truth casewise GRN, the top 10'000 interactions per cell were retained. For each cell-specific network, the AUROC and AUPR were calculated.

## Comparison of RNA velocity methods {#sec:dyngen-velcompare}

@@

## Comparison of trajectory alignment with added noise {#sec:dyngen-tacompare}

2 datasets containing a trajectory were generated using dyngen. For each dataset, we shuffled 5\% to 75\% of the count matrix, with steps of 5\%. Next, we aligned (using DTW) all the base trajectories with all the noisy trajectories.

Let the normalized alignment distance between base trajectory x and trajectory y with z\% noise be $dxy_z$.
Then we looked at the discrimination power of DTW: how well DTW is able to distinguish between two trajectories with a certain amount of added noise. We looked at the following ratios to determine this:
\begin{equation}
    \frac{d12_{z}}{d11_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
\end{equation}

\begin{equation}
    \frac{d21_{z}}{d22_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
\end{equation}

If this ratio is high, it means that DTW is able to distinguish well between the two trajectories. If this ratio is 1, DTW is not able to distinguish between the two trajectories at all.

These datasets and ratios were generated 10 times to reduce possible stochastic effects.

## Code availability {#sec:dyngen-code}

dyngen is available as an open source R package at
\url{https://github.com/dynverse/dyngen}. All code used
in this study is made publicly available at
\url{https://github.com/rcannood/dyngen_analysis}.
